---
title: "Homework3_zc2823"
output: github_document
date: "2025-10-07"
---

```{r setup, include = FALSE}
library(tidyverse)
library(p8105.datasets)
library(knitr)
library(tidyr)
library(forcats)
library(scales)
library(ggplot2)
library(grid)
library(lubridate)
library(janitor)
library(dplyr)
library(patchwork)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.color = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_color_discrete = scale_color_viridis_d
scale_fill_discrete = scale_fill_viridis_d

set.seed(1)
```

# Problem 1

Loading data and read 'instacart'

```{r}
data("instacart")
```

### Size and Structure of the data

```{r}
n_obs <- nrow(instacart)
n_vars <- ncol(instacart)

glimpse(instacart)
```

### Examples of observations

```{r}
example_rows <- instacart |> 
  slice_sample(n = 5)

example_rows
```

#### Short Description of the data

The Instacart dataset contains information on 1,384,617 observations and 15 variables, each corresponding to a single product added to a customer’s online grocery order. Each row represents one product within an order, while columns describe identifiers, product details, and timing information for that purchase, and multiple rows can belong to the same order or user.

Key variables include `order_id` and `user_id` for identifying orders and customers, `product_name` describing the purchased items, `aisle` and `department` indicating the product’s physical or categorical location within the store (for example, yogurt in the dairy eggs department), and `order_dow` and `order_hour_of_day` capturing when the purchases occurred. Also, the variable `reordered` flags whether the product had been previously purchased by the same user.

For example, a sample of five observations shows that user 197159 placed an order (`order_id` 1107870) where `product_id` 33731 was added as the 6th item in the cart and was not ordered by this user in the past.

#### How many aisles are there, and which aisles are the most items ordered from?
Count the number of orders from each aisle

```{r}
aisle_counts <- instacart |>
  count(aisle, name = "n_orders") |>
  arrange(desc(n_orders))
```

Total number of aisles

```{r}
n_aisles <- nrow(aisle_counts)
n_aisles
```

Display the top aisles by number of items ordered

```{r}
head(aisle_counts, 10)
```

##### Answer

There are 134 aisles in the Instacart dataset. Among them, the aisles with the largest number of items ordered are fresh vegetables (150,609 orders) and fresh fruits (150,473 orders), followed by packaged vegetables fruits, yogurt, and packaged cheese. 


#### Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize the plot

Count and keep aisles with > 10,000 items ordered
and Plot

```{r, fig.height=10}
instacart %>%
  count(aisle, name = "items_ordered") %>%
  filter(items_ordered > 10000) %>%
  mutate(aisle = fct_reorder(aisle, items_ordered)) %>%
  ggplot(aes(x = aisle, y = items_ordered)) +
  geom_col(width = 0.7, fill = "steelblue") + 
  coord_flip() +
  scale_y_continuous(
    labels = label_comma(),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    title = "Number of Items Ordered by Aisle (>10,000)",
    x = "Aisle",
    y = "Items Ordered"
  ) +
  theme_minimal(base_size = 9) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.y = element_text(size = 6),
    panel.grid.major.y = element_blank()
  )
```


#### Make a table showing the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”.

Three target aisles

```{r}
target_aisles <- c("baking ingredients", "dog food care", "packaged vegetables fruits")

```

Find top 3 products in each aisle

```{r}
top3_items <- instacart %>%
  filter(aisle %in% target_aisles) %>%
  group_by(aisle, product_name) %>%
  summarize(`Times item ordered` = n(), .groups = "drop_last") %>%
  slice_max(`Times item ordered`, n = 3) %>%
  arrange(aisle, desc(`Times item ordered`))
```

Display as a clean table

```{r}
kable(top3_items, caption = "Top 3 Most Frequently Ordered Products in Selected Aisles")
```


#### Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week

Target two products

```{r}
target_products <- c("Pink Lady Apples", "Coffee Ice Cream")
```

Make labels

```{r}
dow_labels <- c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat")
```

Calculate the mean hour and transform it into a wider format

convert decimal hour to HH:MM format

```{r}
convert_to_time <- function(x) {
  hour <- floor(x)
  minute <- round((x - hour) * 60)
  sprintf("%02d:%02d", hour, minute)
}
```

```{r}
mean_hour_table <- instacart %>%
  filter(product_name %in% target_products) %>%
  mutate(order_dow = factor(order_dow, levels = 0:6, labels = dow_labels)) %>%
  group_by(product_name, order_dow) %>%
  summarize(mean_hour = mean(order_hour_of_day), .groups = "drop") %>%
  mutate(mean_hour_fmt = convert_to_time(mean_hour)) %>%
  select(-mean_hour) %>%
  pivot_wider(
    names_from = order_dow,
    values_from = mean_hour_fmt
  ) %>%
  arrange(product_name)
```

Output a clean table

```{r}
kable(
  mean_hour_table,
  caption = "Mean Order Time of Day (HH:MM) for Selected Items"
)
```


# Problem 2

Import, clean, and tidy the datasets

```{r}
zillow_long <-
  read_csv("data/Zip_zori.csv") |>
  pivot_longer(
    -(RegionID:CountyName),           
    names_to  = "month",
    values_to = "zori"
  ) |>
  janitor::clean_names() |>
  rename(zip_code = region_name) |>
  mutate(
    month    = ymd(month),           
    year     = floor_date(month, "year"),
    zip_code = as.numeric(zip_code)
  ) |>
  select(-county_name) |>
  arrange(zip_code, month)
```

```{r}
zip_meta =
  read_csv("data/Zip_Codes.csv", na = c("NA", ".", "")) |>
  janitor::clean_names() 
```

Make zipcode distinct

```{r}
zip_meta <- zip_meta |> distinct(zip_code, .keep_all = TRUE)
```

Merge

```{r}
zillow_df =
  left_join(zillow_long, zip_meta, by="zip_code") 
```

#### How many ZIP codes are observed 116 times? How many are observed fewer than 10 times? Why are some ZIP codes are observed rarely and others observed in each month?

```{r}
zip_counts <- zillow_long |> 
  group_by(zip_code) |> 
  summarise(n_obs = sum(!is.na(zori)), .groups = "drop")

n_zip_all_116 <- zip_counts |> filter(n_obs == 116) |> nrow()
n_zip_lt_10   <- zip_counts |> filter(n_obs < 10)  |> nrow()
```

Between January 2015 and August 2024 — a total of 116 months — there are `r n_zip_all_116` ZIP codes that are observed in all 116 months, and `r n_zip_lt_10` ZIP codes that are observed fewer than 10 times. 
ZIP codes that appear in every month likely correspond to well-established residential neighborhoods with stable housing markets and consistent Zillow coverage. In contrast, ZIP codes observed rarely (fewer than 10 times) often represent PO Box regions, non-residential or industrial areas, or newly created ZIP codes that Zillow’s dataset does not consistently track. 

#### A table showing the average rental price in each borough and year.

summarize mean ZORI by borough and year and make a table

```{r}
zillow_df |>
  mutate(
    borough = case_match(
      county,
      "Bronx" ~ "Bronx",
      "Kings" ~ "Brooklyn",
      "New York" ~ "Manhattan",
      "Queens" ~ "Queens",
      "Richmond" ~ "Staten Island"
    ),
      year = year(month)) |>
  group_by(borough, year) |>
  summarise(avg_rent = mean(zori, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(
    names_from = year,
    values_from = avg_rent
  ) |>
  mutate(across(where(is.numeric), ~round(.x, 0))) |>
  kable(
    digits = 0,
    caption = "Average Zillow Rental Index (ZORI) by Borough and Year, 2015–2024"
  )
```

Across all five boroughs, average rental prices show a steady upward trend 
from 2015 to 2024. Manhattan consistently has the highest rental prices, 
followed by Brooklyn and Queens, while the Bronx remains the most affordable area. 
After 2020, rental prices in all boroughs appear to rise more sharply, likely 
reflecting post-pandemic recovery and inflation-driven increases in housing demand.

#### Make a plot showing NYC Rental Prices within ZIP codes for all available years

Calculate average rent

```{r}
avg_rent <- 
  zillow_df %>%
  mutate(
    borough = case_match(
      county,
      "Bronx" ~ "Bronx",
      "Kings" ~ "Brooklyn",
      "New York" ~ "Manhattan",
      "Queens" ~ "Queens",
      "Richmond" ~ "Staten Island",
      .default = NA_character_
    ),
    year = year(month)
  ) %>%
  group_by(borough, year) %>%
  summarise(AvgRent = mean(zori, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = borough, values_from = AvgRent)
```

Pivot longer

```{r}
avg_rent_long <- avg_rent %>%
  pivot_longer(
    cols = -year,
    names_to = "borough",
    values_to = "AvgRent"
  )
```

Plot!!!

```{r}
line_chart <- 
ggplot(avg_rent_long, aes(x = year, y = AvgRent, color = borough)) +
  geom_line(linewidth = 1.6) +
  geom_point(size = 2) +
  labs(
    title = "NYC Rental Prices within ZIP Codes for All Available Years",
    x = "Year",
    y = "Average Rental Price ($)",
    color = "Borough"
  ) +
  scale_x_continuous(
    breaks = seq(min(avg_rent_long$year, na.rm = TRUE),
                 max(avg_rent_long$year, na.rm = TRUE),
                 by = 1)
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold")
  )

line_chart
```

The plot shows that rental prices have generally increased across all NYC boroughs from 2015 to 2024. Manhattan remains the most expensive area, followed by Brooklyn and Queens, while the Bronx and Staten Island have lower average rents. A brief dip or plateau around 2020–2021 likely reflects the COVID-19 pandemic’s impact, but prices rose sharply afterward, suggesting a strong post-pandemic recovery in the housing market.

#### Plot showing the distribution of ZIP-code-level rental prices across boroughs

Compute the average rental price within each ZIP code over each month in 2023

```{r}
zip_2023 <-
  zillow_df %>%
  mutate(
    borough = case_match(
      county,
      "Bronx" ~ "Bronx",
      "Kings" ~ "Brooklyn",
      "New York" ~ "Manhattan",
      "Queens" ~ "Queens",
      "Richmond" ~ "Staten Island",
      .default = NA_character_
    ),
    year = year(month)
  ) %>%
  filter(year == 2023, !is.na(borough)) %>%
  group_by(borough, zip_code) %>%
  summarise(avg_rent_2023 = mean(zori, na.rm = TRUE), .groups = "drop")
```

Make box plot

```{r}
box_plot <- 
ggplot(zip_2023, aes(x = borough, y = avg_rent_2023, fill = borough)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.6) +
  labs(
    title = "Distribution of ZIP-Code-Level Rental Prices by Borough (2023)",
    x = "Borough",
    y = "Average Rental Price ($)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(face = "bold")
  )

box_plot
```

The boxplot shows clear differences in ZIP-level rental prices across NYC boroughs in 2023. Manhattan has the highest median rent and the widest range, reflecting both high prices and greater variation within the borough. Brooklyn follows with slightly lower but still elevated rents, while Queens and Bronx sits in the mid-range. The Staten Island have the lowest and most compact distributions, indicating relatively affordable and stable rental markets.

#### Combine the two plots

```{r}
combined_plot <- line_chart / box_plot + 
  plot_annotation(
    title = "NYC Rental Prices: Trends and 2023 Distribution by Borough",
    subtitle = "Top: Average rental price trends (2015–2024) | Bottom: ZIP-level distribution in 2023",
    theme = theme(plot.title = element_text(face = "bold", size = 14))
  )

combined_plot

ggsave(
  filename = "results/nyc_rentals_combined.png",
  plot = combined_plot,
  width = 10,
  height = 10,
  dpi = 300
)
```

# Problem 3

Read demographic and accelerometer data

```{r}
nhanes_covar <- read_csv("data/nhanes_covar.csv", skip = 4)
nhanes_accel <- read_csv("data/nhanes_accel.csv")
```

Merge 

```{r}
nhanes_merged <- nhanes_accel %>%
  left_join(nhanes_covar, by = "SEQN")
```

Exclude participants < 21 years old or with missing demographic data

```{r}
nhanes_clean <- nhanes_merged %>%
  filter(age >= 21) %>%
  drop_na(sex, education, age, BMI)
```

Convert variables to appropriate classes

```{r}
nhanes_clean <- nhanes_clean %>%
  mutate(
    sex = recode_factor(sex, "1" = "Male", "2" = "Female"),
    education = recode_factor(
      education,
      "1" = "Less than high school",
      "2" = "High school or GED",
      "3" = "More than high school",
      .ordered = TRUE
    )
  )

```

#### Table for the number of men and women in each education category

```{r}
edu_sex_table <- nhanes_clean %>%
  group_by(education, sex) %>%
  summarize(`Number of participants` = n(), .groups = "drop") %>%
  pivot_wider(
    names_from = sex,
    values_from = `Number of participants`
  )

kable(
  edu_sex_table,
  caption = "Number of Men and Women by Education Category"
)
```

The table shows that the number of men and women is relatively balanced across education levels, though slightly more participants, both male and female, fall into the “more than high school” category.

#### Visualization of the age distributions for men and women in each education category

```{r}
ggplot(nhanes_clean, aes(x = age, fill = sex)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~education, ncol = 1) +
  theme_minimal() +
  labs(
    title = "Age Distributions by Sex within Each Education Category",
    x = "Age (years)",
    y = "Density",
    fill = "Sex"
  )
```

From the age distribution plots, participants with more than high school education tend to cluster at younger ages, while those with less than high school education show a wider spread extending into older ages. Gender differences in age distribution are minimal overall, although women appear slightly younger on average in the higher education groups.

#### Plot total activities against age 

Calculate total activity over the day

```{r}
nhanes_total <- nhanes_clean %>%
  mutate(
    total_activity = rowSums(across(starts_with("min")), na.rm = TRUE)
  ) %>%
  distinct(SEQN, age, sex, education, total_activity)
```

Visualization

```{r}
ggplot(nhanes_total, aes(x = age, y = total_activity, color = sex)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 0.8) +
  facet_wrap(~ education, nrow = 1) +
  scale_color_viridis_d(option = "plasma", end = 0.8) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title = "Daily Total Physical Activity by Age, Sex, and Education Level",
    x = "Age",
    y = "Total Activity (minutes)",
    color = "sex"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title   = element_text(face = "bold", hjust = 0.5),
    legend.position = "bottom",
    legend.title = element_blank(),
    strip.text   = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.spacing = unit(1, "lines")
  )
```

The figure shows that total daily physical activity tends to decline with age across all education groups.
Participants with less than high school education exhibit the steepest decrease, suggesting that older adults in this group are considerably less active.
Those with more than high school education maintain relatively stable activity levels through middle age, followed by a gradual decline.
Differences between men and women are modest overall. Among younger adults, women show slightly higher total activity on average. At older ages, the two lines converge, indicating comparable activity levels between sexes.
Overall, the plot highlights an inverse relationship between age and daily activity, consistent across education levels but varying in magnitude.

#### Three-panel plot that shows the 24-hour activity time courses for each education level

Pivot longer

```{r}
nhanes_long <- nhanes_clean %>%
  pivot_longer(
    cols = starts_with("min"),
    names_to = "minute",
    values_to = "MIMS"
  ) %>%
  mutate(
    minute = as.numeric(str_remove(minute, "min")),
    hour = floor((minute - 1) / 60)  # 0–23
  )
```

Summarize hourly mean by sex and education

```{r}
hourly_activity <- nhanes_long %>%
  group_by(education, sex, hour) %>%
  summarize(mean_MIMS = mean(MIMS, na.rm = TRUE), .groups = "drop")
```

Plot

```{r}
ggplot(hourly_activity, aes(x = hour, y = mean_MIMS, color = sex)) +
  geom_line(linewidth = 0.5) +
  facet_wrap(~ education, ncol = 1) +
  geom_smooth(se = FALSE, linewidth = 0.6, linetype = "dashed") +
  scale_x_continuous(breaks = seq(0, 24, by = 4)) +
  scale_color_manual(values = c("Male" = "#2B6CB0", "Female" = "#E53E3E")) +
  labs(
    title = "24-hour Activity Patterns by Education Level and Sex",
    x = "Hour of Day",
    y = "Average Activity (mean MIMS)",
    color = "Sex"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

Activity levels are lowest during the early morning hours, increase rapidly after approximately 6–8 AM, and peak around midday (10 AM–2 PM) before gradually declining through the evening.
Across panels, individuals with more than high school education exhibit slightly higher overall activity levels throughout the day, whereas those with less than high school education display lower and more variable activity.
Sex differences are modest: both men and women follow nearly identical daily cycles, although women tend to maintain slightly higher activity levels in the afternoon hours.



