---
title: "Homework3_zc2823"
output: github_document
date: "2025-10-07"
---

```{r setup, include = FALSE}
library(tidyverse)
library(p8105.datasets)
library(knitr)
library(tidyr)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.color = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_color_discrete = scale_color_viridis_d
scale_fill_discrete = scale_fill_viridis_d

set.seed(1)
```

# Problem 1

Loading data and read 'instacart'

```{r}
data("instacart")
```

### Size and Structure of the data

```{r}
n_obs <- nrow(instacart)
n_vars <- ncol(instacart)

glimpse(instacart)
```

### Examples of observations

```{r}
example_rows <- instacart |> 
  slice_sample(n = 5)

example_rows
```

#### Short Description of the data

The Instacart dataset contains information on 1,384,617 observations and 15 variables, each corresponding to a single product added to a customer’s online grocery order. Each row represents one product within an order, while columns describe identifiers, product details, and timing information for that purchase, and multiple rows can belong to the same order or user.

Key variables include `order_id` and `user_id` for identifying orders and customers, `product_name` describing the purchased items, `aisle` and `department` indicating the product’s physical or categorical location within the store (for example, yogurt in the dairy eggs department), and `order_dow` and `order_hour_of_day` capturing when the purchases occurred. Also, the variable `reordered` flags whether the product had been previously purchased by the same user.

For example, a sample of five observations shows that user 197159 placed an order (`order_id` 1107870) where `product_id` 33731 was added as the 6th item in the cart and was not ordered by this user in the past.

#### How many aisles are there, and which aisles are the most items ordered from?
Count the number of orders from each aisle

```{r}
aisle_counts <- instacart |>
  count(aisle, name = "n_orders") |>
  arrange(desc(n_orders))
```

Total number of aisles

```{r}
n_aisles <- nrow(aisle_counts)
n_aisles
```

Display the top aisles by number of items ordered

```{r}
head(aisle_counts, 10)
```

##### Answer

There are 134 aisles in the Instacart dataset. Among them, the aisles with the largest number of items ordered are fresh vegetables (150,609 orders) and fresh fruits (150,473 orders), followed by packaged vegetables fruits, yogurt, and packaged cheese. 


#### Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize the plot

Count and keep aisles with > 10,000 items ordered
and Plot

```{r}
instacart %>%
  group_by(aisle) %>%
  summarize(n = n()) %>%
  filter(n > 10000) %>%
  ggplot(aes(x = reorder(aisle, n), y = n)) +
  geom_point(color = "steelblue", size = 2) +
  scale_y_continuous(
    breaks = seq(0, 150000, by = 20000),   
    labels = scales::comma                 
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    plot.title = element_text(hjust = 0.5, face = "bold")
  ) +
  labs(
    title = "Number of Items Ordered by Aisle (>10,000 Orders)",
    x = "Aisle Name",
    y = "Number of Items Ordered"
  )
```


#### Make a table showing the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”.

Three target aisles

```{r}
target_aisles <- c("baking ingredients", "dog food care", "packaged vegetables fruits")

```

Find top 3 products in each aisle

```{r}
top3_items <- instacart %>%
  filter(aisle %in% target_aisles) %>%
  group_by(aisle, product_name) %>%
  summarize(`Times item ordered` = n(), .groups = "drop_last") %>%
  slice_max(`Times item ordered`, n = 3) %>%
  arrange(aisle, desc(`Times item ordered`))
```

Display as a clean table

```{r}
kable(top3_items, caption = "Top 3 Most Frequently Ordered Products in Selected Aisles")
```


#### Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week

Target two products

```{r}
target_products <- c("Pink Lady Apples", "Coffee Ice Cream")
```

Make labels

```{r}
dow_labels <- c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat")
```

Calculate the mean hour and transform it into a wider format

```{r}
mean_hour_table <- instacart %>%
  filter(product_name %in% target_products) %>%
  mutate(order_dow = factor(order_dow, levels = 0:6, labels = dow_labels)) %>%
  group_by(product_name, order_dow) %>%
  summarize(`Mean order hour` = mean(order_hour_of_day), .groups = "drop") %>%
  pivot_wider(
    names_from = order_dow,
    values_from = `Mean order hour`
  ) %>%
  arrange(product_name)
```

Output a clean table

```{r}
kable(mean_hour_table, digits = 2,
      caption = "Mean Hour of the Day When Selected Items Are Ordered")
```






